Viscous Transport in Eroding Porous Media

### 2019-08-21 docs/transport/erosion_transport.tex
finished a very rough first draft of the intro

In many applications, the fluid transport continuously changes the pore
structure.  This is particularly common in geophysical flows where
erosion alters the porous matrix.  In this work, we investigate flow in
eroded porous media.  By introducing erosion, there is a two-way
coupling between the hydrodynamics and the mechanical alteration of the
complex porous geometry.  Transport in porous geometries that have
undergone erosion finds applications in aquifers~\cite{cve-che-wen1996,
sud1986}, solute transport in groundwater flow~\cite{dag1987,
kon-bre1978}.  Porous geometries that have undergone erosion have
certain characteristics.  For example, the geometry often has regions of
high porosity that connect the inlet to the outlet.  This channelization
results in velocity scales that can vary over several orders of
magnitude~\cite{all-hea-lab-rei2002}.  Moreover, the channelization
results in a geometry that is largely heterogeneous and anisotropic.
Others have studied transport in heterogeneous geometries (for example,
see~\cite{mil-chr-imh-mcb-ped1998, ber-sch-sil2000, hak-com-den2019,
suo-liu-gan2019, cus-hu-den1995, cve-che-wen1996, leb-ded-dav-bou2007,
knu-car2005, bel-sal-rin1992}). However, to the best of our knowledge, a
detailed study of transport in eroded geometries has not been performed.

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
hydrodynamics, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018, par-izu2000}.  A variety of
numerical methods can be applied to solve the incompressible Stokes
equations such as a lattice-gas model~\cite{kop-kat-tim1996}, finite
volume~\cite{suo-liu-gan2019, den-ica-hid2018, sie-ili-pri-riv-gua2019},
finite differences~\cite{leb-ded-dav-bou2007, knu-car2005}, and cellular
automata~\cite{rot1988}. In this work, we instead solve an equivalent
boundary integral equation (BIE) formulation of the incompressible
Stokes equation.  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.  Given the solution of the incompressible Stokes equations are
solved in an eroded porous geometry, we investigate different metrics to
characterize transport.  We focus on three metrics: the tortuosity, the
anomalous dispersion, and the distribution of geometrical channels.

@@

The numerical methods to solve the governing fluid and erosion equations
are based on previous work of the authors~\cite{qua-moo2018}.  We
continue to use a boundary integral equation (BIE) formulation of the
incompressible Stokes equations and a second-order Runge-Kutta method
applied to a smoothed version of the magnitude of the shear stress.  In
our previous work, we discretized the BIE with the trapezoid rule whose
accuracy grows when evaluating layer potentials close to the fluid
boundary.  Therefore, given a fixed resolution, our previous work
required a minimum separation distance between all bodies to maintain
stability.  In recent years, many near-singular integration schemes have
been developed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a,
bea-yin-wil2016, bea-lai2001, klo-bar-gre-one2013}.  In this work, we
use a Barycentric quadrature method~\cite{bar2014, bar-wu-vee2015} since
it is a non-intrusive modification of the trapezoid rule, and the error
is guaranteed to be uniformly bounded.  In this work, we extend this
quadrature method so that it can be used to compute the gradient of the
velocity which is required to compute the shear stress and the fluid
vorticity.

@@

\paragraph{Related Work}

\begin{itemize}
\item \cite{dar-mcc1998} show that if there is a region with high
porosity, this greatly affects the relationships between the porosity
and permeability.  Uses a Lattice-Boltzmann method.

\item Rothman in 1988 did CA for flow in 2D porous media

\item ``Canductivity and permeability af rocks" 1984

\item ``An Analytical Study on Natural Convection in Isotropic and .
Anisotropic Porous Channels" shows how anisotropy affects heat transfer
in porous media.

\item "The Effect of Anisotropic Permeability on Free Convective
Boundary Layer Flow in Porous Media" is the same as last bullet

\item Thermal boundary layer flow in anisotropic porous media is
investigated by Nilsen and Storesletten~\cite{nil-sto1990} and Rees and
Storesletten~\cite{ree-sto1995}.

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\cite{bak-she1986}

We start by defining the main variables used to model erosion.  A more
detailed description is in our previous work~\cite{qua-moo2018}.  We
consider flows inside a confined geometry $\Omega$ that contains $M$
eroding bodies with boundary $\gamma_\ell$.  The boundary of the
bounding geometry is $\Gamma$, and the fluid domain is bounded by $\bd
\Omega = \Gamma \cup \gamma_1 \cup \cdots \cup \gamma_M$.  On each
eroding body, $\nn$ is the normal vector pointing into the body, and
$\ss$ is the unit tangent vector pointing in the counterclockwise
direction.  The normal vector on $\Gamma$ is the outward unit normal.
Neglecting inertial forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}\\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass} \\
    \uu = 0, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the bodies} \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity} \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Using the same approach as
our previous work~\cite{qua-moo2018}, we use a double-layer potential
formulation for the velocity $\uu$.  The double-layer potential is
\begin{align}
  \DDD[\eeta](\xx) = \int_{\bd\Omega} D(\xx,\yy) \eeta(\yy)\, ds_\yy = 
  \frac{1}{\pi}\int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
  \label{eqn:stokesDLP}
\end{align}
where $D$ is the kernel of the integral operator, $\rr = \xx - \yy$,
$\rho = \|\rr\|$, $\nn$ is the unit outward normal at $\yy$, and $\eeta$
is an unknown density function.  Since the Stokes double-layer potential
can not represent all solutions of the incompressible Stokes equations,
we complete the integral equation formulation by adding the Stokeslets
and rotlets~\cite{pow-mir1987}
\begin{align}
  S[\llambda_\ell,\cc_\ell](\xx) = \frac{1}{4\pi} \left( 
    -\log \rho \II + \frac{\rr \otimes \rr}{\rho^2} \right), \quad
  R[\xi_\ell,\cc_\ell](\xx) = \frac{\rr^\perp}{\rho^2} \xi_\ell,
\end{align}
where $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ eroding body, $\rr^\perp = (r_2,-r_1)$, and $\rho =
\|\rr\|$.  Then, for any sufficiently smooth geometry $\Omega$, the
solution of the incompressible Stokes equation with a Dirichlet boundary
condition $\ff$ is
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx),
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx) +
    \NN_0[\eeta](\xx), \quad \xx \in \bd\Omega, \\
  \llambda_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell} 
    \eeta(\yy)\, ds_\yy \quad \ell = 1,\ldots,M, \\
  \xi_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell}
    (\yy - \cc_\ell)^\perp \cdot \eeta(\yy)\, ds_\yy 
    \quad \ell = 1,\ldots,M.
\end{align}
For our particular problem, $\ff$ is the prescribed velocity on the
outer wall and is zero on the eroding bodies.  The operator $\NN_0$ is
the integral operator with kernel $\nn(\xx) \otimes \nn(\yy)$, $\xx,\yy
\in \Gamma$, which removes the rank one null space resulting from the
flux-free requirement of $\ff$.

>>>>

In many applications, the fluid transport continuously changes the pore
structure.  This is particularly common in geophysical flows where
erosion alters the porous matrix.  In this work, we investigate flow in
eroded porous media.  By introducing erosion, there is a two-way
coupling between the hydrodynamics and the mechanical alteration of the
complex porous geometry.  Transport in porous geometries that have
undergone erosion finds applications in aquifers~\cite{cve-che-wen1996,
sud1986}, solute transport in groundwater flow~\cite{dag1987,
kon-bre1978}.  Porous geometries that have undergone erosion have
certain characteristics.  For example, the geometry often has regions of
high porosity that connect the inlet to the outlet.  This channelization
results in velocity scales that can vary over several orders of
magnitude~\cite{all-hea-lab-rei2002}.  Moreover, the channelization
results in a geometry that is largely heterogeneous and anisotropic, and
this affects heat transfer through the medium~\cite{nil-sto1990,
ree-sto1995}.  Others have studied transport in heterogeneous geometries
(for example, see~\cite{mil-chr-imh-mcb-ped1998, ber-sch-sil2000,
hak-com-den2019, suo-liu-gan2019, cus-hu-den1995, cve-che-wen1996,
leb-ded-dav-bou2007, knu-car2005, bel-sal-rin1992}). However, to the
best of our knowledge, a detailed study of transport in eroded
geometries has not been performed.

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
hydrodynamics, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018, par-izu2000}.  A variety of
numerical methods can be applied to solve the incompressible Stokes
equations such as a lattice-gas or lattice Boltzmann
model~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volume~\cite{suo-liu-gan2019, den-ica-hid2018, sie-ili-pri-riv-gua2019},
finite differences~\cite{leb-ded-dav-bou2007, knu-car2005}, and cellular
automata~\cite{rot1988}. In this work, we instead solve an equivalent
boundary integral equation (BIE) formulation of the incompressible
Stokes equation.  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.  Given the solution of the incompressible Stokes equations are
solved in an eroded porous geometry, we investigate different metrics to
characterize transport.  We focus on three metrics: the tortuosity, the
anomalous dispersion, and the distribution of geometrical channels.

@@

The numerical methods to solve the governing fluid and erosion equations
are based on previous work of the authors~\cite{qua-moo2018}.  We
continue to use a boundary integral equation (BIE) formulation of the
incompressible Stokes equations and a second-order Runge-Kutta method
applied to a smoothed version of the magnitude of the shear stress.  In
our previous work, we discretized the BIE with the trapezoid rule whose
accuracy grows when evaluating layer potentials close to the fluid
boundary.  Therefore, given a fixed resolution, our previous work
required a minimum separation distance between all bodies to maintain
stability.  One of the earliest works on near-singular integration was
developed by Baker and Shelley~\cite{bak-she1986}, and in recent years,
many other near-singular integration schemes have been
developed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  In this work, we use a Barycentric
quadrature method~\cite{bar2014, bar-wu-vee2015} since it is a
non-intrusive modification of the trapezoid rule, and the error is
guaranteed to be uniformly bounded.  In this work, we extend this
quadrature method so that it can be used to compute the gradient of the
velocity which is required to compute the shear stress and the fluid
vorticity.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we briefly summarize the model that we
describe in detail in our previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:method} describes the numerical methods, with special
attention paid to the new quadrature method to compute the shear stress
and vorticity.  Section~\ref{sec:transport} describes measures for
characterizing the geometry, transport, and dispersion.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  A more
detailed description is in our previous work~\cite{qua-moo2018}.  We
consider flows inside a confined geometry $\Omega$ that contains $M$
eroding bodies with boundary $\gamma_\ell$.  The boundary of the
bounding geometry is $\Gamma$, and the fluid domain is bounded by $\bd
\Omega = \Gamma \cup \gamma_1 \cup \cdots \cup \gamma_M$.  On each
eroding body, $\nn$ is the normal vector pointing into the body, and
$\ss$ is the unit tangent vector pointing in the counterclockwise
direction.  The normal vector on $\Gamma$ is the outward unit normal.
Neglecting inertial forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}\\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass} \\
    \uu = 0, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the bodies} \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity} \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Using the same approach as
our previous work~\cite{qua-moo2018}, we use a double-layer potential
formulation for the velocity $\uu$.  The double-layer potential is
\begin{align}
  \DDD[\eeta](\xx) = \int_{\bd\Omega} D(\xx,\yy) \eeta(\yy)\, ds_\yy = 
  \frac{1}{\pi}\int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
  \label{eqn:stokesDLP}
\end{align}
where $D$ is the kernel of the integral operator, $\rr = \xx - \yy$,
$\rho = \|\rr\|$, $\nn$ is the unit outward normal at $\yy$, and $\eeta$
is an unknown density function.  Since the Stokes double-layer potential
can not represent all solutions of the incompressible Stokes equations,
we complete the integral equation formulation by adding the Stokeslets
and rotlets~\cite{pow-mir1987}
\begin{align}
  S[\llambda_\ell,\cc_\ell](\xx) = \frac{1}{4\pi} \left( 
    -\log \rho \II + \frac{\rr \otimes \rr}{\rho^2} \right), \quad
  R[\xi_\ell,\cc_\ell](\xx) = \frac{\rr^\perp}{\rho^2} \xi_\ell,
\end{align}
where $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ eroding body, $\rr^\perp = (r_2,-r_1)$, and $\rho =
\|\rr\|$.  Then, for any sufficiently smooth geometry $\Omega$, the
solution of the incompressible Stokes equation with a Dirichlet boundary
condition $\ff$ is
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx),
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx) +
    \NN_0[\eeta](\xx), \quad \xx \in \bd\Omega, \\
  \llambda_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell} 
    \eeta(\yy)\, ds_\yy \quad \ell = 1,\ldots,M, \\
  \xi_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell}
    (\yy - \cc_\ell)^\perp \cdot \eeta(\yy)\, ds_\yy 
    \quad \ell = 1,\ldots,M.
\end{align}
For our particular problem, $\ff$ is the prescribed velocity on the
outer wall and is zero on the eroding bodies.  The operator $\NN_0$ is
the integral operator with kernel $\nn(\xx) \otimes \nn(\yy)$, $\xx,\yy
\in \Gamma$, which removes the rank one null space resulting from the
flux-free requirement of $\ff$.
